cardinalR: Generating interesting high-dimensional data structures

### 2025-04-26 paper-cardinalR.Rmd
Added

```{r main-tb-pdf, eval=knitr::is_latex_output()}
main_tb |> 
  kable(caption = "The main arguments for `gen_multicluster()`.", format="latex", col.names = c("Argument", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

>>>>

```{r main-tb-pdf, eval=knitr::is_latex_output()}
main_tb |> 
  kable(caption = "The main arguments for gen\\_multicluster().", format="latex", col.names = c("Argument", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

### 2025-04-26 paper-cardinalR.tex
Added

\caption{\label{tab:main-tb-pdf}The main arguments for `gen_multicluster()`.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Argument & Explanation\\
\midrule
n & A numeric vector representing the sample sizes.\\
p & A numeric value representing the number of dimensions.\\
k & A numeric value representing the number of clusters.\\
loc & A numeric matrix representing the locations/centroids of clusters.\\
scale & A numeric vector representing the scaling factors of clusters.\\
\addlinespace
shape & A character vector representing the shapes of clusters.\\
rotation & A numeric list which contains plane and the corresponding angle along that plane for each cluster.\\
is\_bkg & A Boolean value representing the background noise should exist or not.\\
\bottomrule
\end{tabular}
\end{table}

@@

We illustrate the use of the package by generating five clusters of different shapes, positioned with equal inter-cluster distances in \(4\text{-}D\). Using this synthetic dataset, we will evaluate the performance of six popular dimension reduction techniques: Principal Component Analysis (PCA) \citep{jolliffe2011}, t-distributed stochastic neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and projection (UMAP) \citep{leland2018}, potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019}, large-scale dimensionality reduction Using triplets (TriMAP) \citep{amid2019}, and pairwise controlled manifold approximation (PaCMAP) \citep{yingfan2021}.

\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/unnamed-chunk-13-1}

>>>>

\caption{\label{tab:main-tb-pdf}The main arguments for gen\_multicluster().}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Argument & Explanation\\
\midrule
n & A numeric vector representing the sample sizes.\\
p & A numeric value representing the number of dimensions.\\
k & A numeric value representing the number of clusters.\\
loc & A numeric matrix representing the locations/centroids of clusters.\\
scale & A numeric vector representing the scaling factors of clusters.\\
\addlinespace
shape & A character vector representing the shapes of clusters.\\
rotation & A numeric list which contains plane and the corresponding angle along that plane for each cluster.\\
is\_bkg & A Boolean value representing the background noise should exist or not.\\
\bottomrule
\end{tabular}
\end{table}

@@

This section illustrates the use of package by generating a synthetic dataset to evaluate the performance of six popular dimension reduction techniques: Principal Component Analysis (PCA) \citep{jolliffe2011}, t-distributed stochastic neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and projection (UMAP) \citep{leland2018}, potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019}, large-scale dimensionality reduction Using triplets (TriMAP) \citep{amid2019}, and pairwise controlled manifold approximation (PaCMAP) \citep{yingfan2021}.

We generated a dataset of five clusters, positioned with equal inter-cluster distances in \(4\text{-}D\) space. The five clusters have different geometric structures and each contain different number of points. Specifically, the helical spiral cluster includes \(2250\) points and was generated with a scale parameter of \(0.4\). The hemisphere cluster consists of \(1500\) points with a scale parameter of \(0.35\). The uniform cube-shaped cluster contains \(750\) points and uses a scale parameter of \(0.3\). The blunted cone cluster includes \(1250\) points, generated with a scale parameter of \(1\). Finally, the Gaussian-shaped cluster contains \(1750\) points and was generated with a scale parameter of \(0.3\).

\begin{verbatim}
positions <- geozoo::simplex(p=4)$points
positions <- positions * 0.8

## To generate data
five_clusts <- gen_multicluster(n = c(2250, 1500, 750, 1250, 1750), p = 4, k = 5,
                       loc = positions,
                       scale = c(0.4, 0.35, 0.3, 1, 0.3),
                       shape = c("helicalspiral", "hemisphere", "unifcube", 
                                 "bluntedcone", "gaussian"),
                       rotation = NULL,
                       is_bkg = FALSE)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-nldr-layouts-1} \caption{Six Drs}\label{fig:fig-nldr-layouts}
\end{figure}

UMAP, PHATE, TriMAP, and PaCMAP effectively separate the five clusters and show the preservation of the global structure. However, PHATE reveals three non-linear clusters, even though two of them do not show non-linearity. UMAP, TriMAP, and PaCMAP successfully maintain the local structures of the data. In contrast, tSNE divides the non-linear cluster into sub-clusters. Also, tSNE fails to preserve the distances between the clusters. PCA, on the other hand, preserves the local structures of the clusters, but some clusters are incorrectly merged that should remain distinct.
